\chapter{Introduction}

%It is \textit{Year 2 after Snowden} and we lost trust in every single component of digital communication. In every single one? No! A tiny standard resists the pressure and keeps our secrets what they should be, secret. I proudly introduce: OpenPGP! \\


On insecure networks like the Internet it is possible for everyone eavesdropping on the communication to read the transmitted message. In addition it is possible to change the content of such a message. Furthermore an attacker can spoof a message, claiming the identity of a third party.
To securely exchange messages using insecure networks like the Internet it is therefore required to protect the transmitted message. The favored properties of such a protection are thus confidentiality, integrity and authenticity. \\


% 3 ... openpgp
The \textit{OpenPGP Message Protocol} is an Internet standard providing exactly those properties.  This is done by combining several algorithms to encrypt messages ahead of transmitting them. In addition digital signatures are used to provide integrity. OpenPGP provides an end-to-end protection of messages. It is therefore possible to use untrusted communication components for transmission.
In chapter \ref{chapter:openpgp} of this thesis we give an overview of the OpenPGP Message Protocol and its ecosystem. \\

% 4 ... message format / data / algorithms
OpenPGP achieves the properties mentioned by defining a structure for data and various procedures. The details of OpenPGP's data structures and how they are used by the algorithms are discussed in chapter \ref{chapter:messageformat}. \\


% 5 ... jpg
We used the \myacro{IAIK-JCE} toolkit for the Java programming language in our own implementation of OpenPGP. Doing so we implemented the data structures defined in the OpenPGP standard. Furthermore we used the IAIK-JCE to provide the required procedures specified by the OpenPGP standard. We implemented encryption, signing and key management. Our implementation is based on the Java stream architecture. We present this implementation, the \textit{Java Privacy Guard}, in chapter \ref{chapter:jpg}. \\

% 6 ... concerns
One concern of the OpenPGP standard affects the used algorithms. Since various algorithms are used in OpenPGP its security depends on the security of those algorithms. In addition it is required to carefully pick values for the parameters of those algorithms. Another concern applies to the usability of OpenPGP implementations. Since OpenPGP is used in a non-transparent way, it requires a lot of user interaction.
Chapter \ref{chapter:concerns} covers general security and usability concerns.

%
%The introduction is your entire thesis in, say, 1-2 pages. If someone reads
%the intro, she should have understood your thesis. Maybe not all the
%details, but nothing that comes afterwards will be a surprise.
%
%This chapter is the second thing you write. Needs a major overhaul
%at the end. Why do you write Abstract and Introduction first?
%Because it forces you to think about the structure of the work and
%the reasons for doing what you did. There is nothing worse than
%writing four chapters and then figuring out you did not understand
%the rationale for your work --- you would have to rewrite a lot.
%
%The structure of the introduction is as follows:
%
%1. Background. What are you talking about? A new methods for public
%key cryptography? Then talk about communication security. A new
%algorithm for image recognition? Talk about the importance of image
%recognition is modern life.
%
%2. The problem. Something was wrong in the world when you started
%writing your thesis. What was it? Why was it a problem? Here you
%state the problem and set concrete goals for a solution. For
%instance, people with quantum computers could break everyone's
%communication. OR we have a new algorithm, but we are not sure if
%it is secure. This needs to be solved to make the world a better
%place.
%
%3. Your solution. How did you address the problem? Describe what
%you did, and in how far it solves the problem. Describe drawbacks
%as well, but they should be secondary. Your contribution could be
%that now we know that the given algorithm is insecure. This being
%a Master's Thesis, you are not expected to fulfill all the goals you
%set.
%
%4.Structure of thesis. What are you going to describe in which
%chapter?
%
%Again, the introduction is important. By the end of it, you have stated
%that your field of research is important, that there was a problem and
%that you have solved some aspect of it. In some sense, it is a sales
%pitch,
%but a sales pitch that does not exagerate anything.
%
%Throughout this chapter and the other chapters, you will sprinkle
%references to related work to show that you know what is going on
%beside your work. You use references to back up claims
%("X.509 is generally considered to be faulty
%[ImportantResearcher13]"), but also to give the other side of the
%debate without spending too many words ("although the importance is
%privacy in the facebook age is under discussion [Zuc12]").
%
%Under no circumstance do you claim something without backing it up.
%If you have a wild personal claim that you cannot back up ("SSL is
%broken") you use it as a working assumption ("In this thesis, we
%will examine alternatives to the SSL architecture that are immune
%to the possible criticism that ... .") A favorite trick of the
%trade is to downplay a claim. For instance, instead of saying that
%something is insecure, you can say that its insecurity has been
%argued by someone, or even that security is not completely obvious.
%The reader understands what you are really saying, and for practical
%purposes, these claims mean about the same. (But don't underclaim
%either!)